Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimization Choices

### 2025-10-24 revision/response.qmd

7. I was also wondering how the dependence of the smoothness on d can be explained, since it is not intuitive that this should change so much? (Note also that the last sentence in the caption for Table 1 seems to mix up squintability and smoothness.)

**\textcolor{purple}{TODO The caption has been fixed.}**

8. Finally, I appreciate the practical advice in section 7. Here, 7.1 should be somewhat extended in my opinion: does the planned tour then run through all the individual paths, should we follow the most successful jellyfish, or how would we pick?

**\textcolor{purple}{We make Section 7.1 clearer that the planned tour can be run to trace the optimisation of individual jellyfish path. Users may wish to follow the most successful jellyfish, a random selection, or interesting jellyfish that result in local optimum. DISCUSSION: should I include a small example with the code?}**

@@

Finally I have also collected a few minor remarks:

  * After defining Eq. 4 the additional parameter should probably be theta_4 and not theta_2.  **\textcolor{purple}{our responses are in red.}**
  * In the algorithm pseudo-code it is not clear what c_t and the trend actually are.  **\textcolor{purple}{In the text and the pseudo-code, we include information that $c_t$ is the time control function determine the exploration and exploitation phase of the algorithm. We also include information that the (ocean) trend is defined as the difference between the best jellyfish and the average of all current jellyfish.}**
  * In Fig. 4 please specify what optimizer was used. **\textcolor{purple}{We have included the information in the caption that the CRS optimised is used for this illustration. }**
  * From the last sentence of Section 5, it is not clear to me why the ranks need to be used, why not scale the metrics?  **\textcolor{purple}{our responses are in red}**.
  * In 6.2 it is not really clear what is being bootstrapped for the evaluation of the uncertainty. **\textcolor{purple}{We have included further description in the paper to describe the bootstrap procedure as drawing 500 bootstrap samples with replacement from the original 50 simulations for each setting: an optimisation with high-dimensional data (d = 4, 6, 8, 10, or 12) with different number of jellyfish (20, 50, or 100) and maximum number of iteration (50 or 100)}**.

### 2025-10-24 jso.qmd
more revision

To solve the optimization problem embedded in the PPGT with JSO, a starting
projection, an index function, the number of jellyfish, and the maximum
number of iterations are provided as input. Then, the current
projection is evaluated by the index function. The projection is then
moved in a direction determined by a random factor, influenced by how
far along we are in the optimization process. Occasionally, completely
new directions may be taken like a jellyfish might with ocean currents.
A new projection is accepted if it is an improvement compared to the
current one, rejected otherwise. This process continues and iteratively
improves the projection, until the pre-specified maximum number of
trials is reached.

@@

> **if** $c_t$ is greater than or equal to $0.5$ **then**
>
> > Define trend based on the `current_best` and `current_projections`
>
> > Update each projection towards the trend using a random factor and
> > orthonormalisation
>
> **else**
>
> > **if** a random number is greater than $1 - c_t$ **then**
> >
> > > Slightly adjust each projection with a small random factor
> > > (passive)
> >
> > **else**
> >
> > > For each projection, compare with a random jellyfish in `current_projections`
> > > and adjust towards or away from it based on their corresponding `current_idx` (active)
>
> Update the orientation of each projection to maintain consistency
>
> Evaluate the new projections using the index function

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters
that are most relevant to our application: the number of jellyfish and
the maximum number of iterations

@@

```{r}
#| label: fig-success-rate
#| fig.cap: "How success rate is calculated, illustrated using the optimal projections from 50 optimizations of 8D pipe data sorted by index value. The pipe shape is recognizable in the projection index values between 0.933-0.969. Of the 50 simulations, 43 achieved an index value within 0.05 of the best, resulting in a success rate of 0.86."
# pipe_jellyfish_12d <- pipe_better |> 
#   filter(dim == 2) |> 
#   mutate(dim = as.numeric(dim) * 8, sim = as.numeric(sim)) 
# proj_dt <- map_dfr(pipe_jellyfish_12d |> pull(basis),
#               ~as_tibble(pipe1000_8d %*% .x), .id = "sim") |>
#   mutate(sim = as.numeric(sim)) |>
#   left_join(pipe_jellyfish_12d |> select(sim, index_val), by = "sim")
# 
# idx_val_sorted <- proj_dt |> pull(index_val) |> unique() |> sort()
# proj_dt |>
#   ggplot(aes(x = V1, y = V2)) +
#   geom_point(size = 0.5) +
#   geom_text(aes(label = sprintf("%.3f", index_val)), x = 0, y = 3.6,
#             size = 5) +
#   xlim(-4, 4) + ylim(-4, 4) +
#   facet_wrap(~fct_reorder(as.factor(sim), -index_val), ncol = 10) +
#   theme_bw() +
#   theme(aspect.ratio = 1, axis.ticks = element_blank(),
#         axis.text = element_blank(), axis.title = element_blank(),
#         panel.grid = element_blank(), strip.background = element_blank(),
#         strip.text = element_blank()) 
knitr::include_graphics(here::here("figures", "success-rate.png"))
```

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of iteration) on the success rate is summarized in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates sampled from 
the simulation for each case. As the number of jellyfish and maximum iteration allowed 
increase, the success rate also increases. For problems in lower dimensional (4,
6) search spaces, small parameter values (20 jellyfish and 50 iterations) 
are sufficient to achieve a high success rate. Many more
jellyfish (100 jellyfish and 100 iterations) are needed for finding the 
pipe structure in d = 10, and 12. This suggests
that even JSO will struggle with high dimensions, and require
substantially more computational time to have an acceptable success rate,
especially for difficult PP indexes based on scagnostics.

@@

To use JSO for PPGT in the `tourr` package, specify
`search_f = search_jellyfish` in the guided tour.

The animation function `animate_*()` won't directly render the animated 
projection for JSO, as with other existing optimizers since multiple tour 
paths will need to be generated. To
visualize the path visited by each individual jellyfish, assign the
animation to an object (`res <- animate_xy(...)`). This will save the
bases visited by JSO, along with relevant metadata, as a tibble data
object [see @RJ-2021-105 for more details on the data object]. The bases
visited for individual jellyfish can then be extracted and viewed using
the `planned_tour()` function.

>>>>

To solve the optimization problem embedded in the PPGT with JSO, a starting
projection, an index function, the number of jellyfish, and the maximum
number of iterations are provided as input. Then, the current
projection is evaluated by the index function. The projection is then
moved in a direction determined by a random time control function, $c_t$, whose value decreases as 
the number of trials increases (subject to randomness) to guide the exploration and exploitation phases 
of the algorithm. When $c_t \ge 0.5$, new 
directions will be taken like a jellyfish explores with the ocean trend, defined as 
the difference of the current best jellyfish and average of all the current jellyfish.
When $1- c_t$ is smaller than a randomly generated number in $[0, 1]$, jellyfish will move passively with small random perturbations, otherwise, it will move actively towards or away from another jellyfish based on their index values. 
A new projection is accepted if it is an improvement compared to the
current one, rejected otherwise. This process continues and iteratively
improves the projection, until the pre-specified maximum number of
trials is reached.

@@

> **if** $c_t$ is greater than or equal to $0.5$ **then**
>
> > Define the ocean current trend as the difference of the `current_best` and the average of the `current_projections`
>
> > Update each projection towards the ocean trend using a random factor and
> > orthonormalisation
>
> **else**
>
> > **if** a random number is greater than $1 - c_t$ **then**
> >
> > > Slightly adjust each projection with a small random factor
> > > (passive)
> >
> > **else**
> >
> > > For each projection, compare with a random jellyfish in `current_projections`
> > > and adjust towards or away from it based on their corresponding `current_idx` (active)
>
> Update the orientation of each projection to maintain consistency
>
> Evaluate the new projections using the index function

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. While the specific implementation
details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters
that are most relevant to our application: the number of jellyfish and
the maximum number of iterations

@@

```{r}
#| label: fig-success-rate
#| fig.cap: "How success rate is calculated, illustrated using the optimal projections from 50 optimizations of 8D pipe data, optimised by CRS, sorted by index value. The pipe shape is recognizable in the projection index values between 0.933-0.969. Of the 50 simulations, 43 achieved an index value within 0.05 of the best, resulting in a success rate of 0.86."
# pipe_jellyfish_12d <- pipe_better |> 
#   filter(dim == 2) |> 
#   mutate(dim = as.numeric(dim) * 8, sim = as.numeric(sim)) 
# proj_dt <- map_dfr(pipe_jellyfish_12d |> pull(basis),
#               ~as_tibble(pipe1000_8d %*% .x), .id = "sim") |>
#   mutate(sim = as.numeric(sim)) |>
#   left_join(pipe_jellyfish_12d |> select(sim, index_val), by = "sim")
# 
# idx_val_sorted <- proj_dt |> pull(index_val) |> unique() |> sort()
# proj_dt |>
#   ggplot(aes(x = V1, y = V2)) +
#   geom_point(size = 0.5) +
#   geom_text(aes(label = sprintf("%.3f", index_val)), x = 0, y = 3.6,
#             size = 5) +
#   xlim(-4, 4) + ylim(-4, 4) +
#   facet_wrap(~fct_reorder(as.factor(sim), -index_val), ncol = 10) +
#   theme_bw() +
#   theme(aspect.ratio = 1, axis.ticks = element_blank(),
#         axis.text = element_blank(), axis.title = element_blank(),
#         panel.grid = element_blank(), strip.background = element_blank(),
#         strip.text = element_blank()) 
knitr::include_graphics(here::here("figures", "success-rate.png"))
```

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of iteration) on the success rate is summarized in @fig-proportion.
Bootstrap resampling is used to quantify the uncertainty of the success rate through 500 bootstrap samples with replacement from the original 50 simulation results for each task: optimizing a high-dimensional data (d = 4, 6, 8, 10, or 12) with different number of jellyfish ( 20, 50, or 100) and maximum number of iteration (50 or 100). 
As the number of jellyfish and maximum iteration allowed 
increase, the success rate also increases. For problems in lower dimensional (4,
6) search spaces, small parameter values (20 jellyfish and 50 iterations) 
are sufficient to achieve a high success rate. Many more
jellyfish (100 jellyfish and 100 iterations) are needed for finding the 
pipe structure in d = 10, and 12. This suggests
that even JSO will struggle with high dimensions, and require
substantially more computational time to have an acceptable success rate,
especially for difficult PP indexes based on scagnostics.

@@

To use JSO for PPGT in the `tourr` package, specify
`search_f = search_jellyfish` in the guided tour. 
The animation function `animate_*()` won't directly render the animated 
projection for JSO, as with other existing optimizers since multiple tour 
paths will need to be generated. To
visualize the path visited by each individual jellyfish, assign the
animation to an object (`res <- animate_xy(...)`). This will save all the
bases visited by JSO, along with relevant metadata, as a tibble data
object [see @RJ-2021-105 for more details on the data object]. Users can 
then view the path of individual jellyfish through extracting the relevant 
bases recorded and construct it as a planned tour (using the `planned_tour()` function).
Users may wish to view the 
optimisation path of the most successful jellyfish, or a random selection of jellyfish, 
or some specific runs that find local maximum.

### 2025-10-24 jso.tex
more revision

It is expected that for a PP index with high squint angle, the
optimization \eqref{eq-optimization} should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation, shown in the
simulated paths from different index functions in
Figure~\ref{fig-squintability-ill}, suggests that the half-point index
value improvement percentage provides an appropriate mathematical
definition of squintability, which matches the intuition behind this
concept, while being amenable to numerical computation.

@@

To solve the optimization problem embedded in the PPGT with JSO, a
starting projection, an index function, the number of jellyfish, and the
maximum number of iterations are provided as input. Then, the current
projection is evaluated by the index function. The projection is then
moved in a direction determined by a random factor, influenced by how
far along we are in the optimization process. Occasionally, completely
new directions may be taken like a jellyfish might with ocean currents.
A new projection is accepted if it is an improvement compared to the
current one, rejected otherwise. This process continues and iteratively
improves the projection, until the pre-specified maximum number of
trials is reached.

\begin{tcolorbox}[enhanced jigsaw, breakable, opacityback=0, toprule=.15mm, toptitle=1mm, titlerule=0mm, left=2mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, coltitle=black, bottomrule=.15mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm]

@@

\begin{quote}
Define trend based on the \texttt{current\_best} and
\texttt{current\_projections}
\end{quote}

\begin{quote}
Update each projection towards the trend using a random factor and
orthonormalisation
\end{quote}

@@

\caption{\label{fig-success-rate}How success rate is calculated,
illustrated using the optimal projections from 50 optimizations of 8D
pipe data sorted by index value. The pipe shape is recognizable in the
projection index values between 0.933-0.969. Of the 50 simulations, 43
achieved an index value within 0.05 of the best, resulting in a success
rate of 0.86.}

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of iteration) on the success rate is summarized in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates sampled from the simulation for each case. As the
number of jellyfish and maximum iteration allowed increase, the success
rate also increases. For problems in lower dimensional (4, 6) search
spaces, small parameter values (20 jellyfish and 50 iterations) are
sufficient to achieve a high success rate. Many more jellyfish (100
jellyfish and 100 iterations) are needed for finding the pipe structure
in d = 10, and 12. This suggests that even JSO will struggle with high
dimensions, and require substantially more computational time to have an
acceptable success rate, especially for difficult PP indexes based on
scagnostics.

@@

\caption{\label{tbl-smoothness-squintability}The columns \(\nu\) and
\(\varsigma\) show the smoothness and squintability measures for three
PP indexes for the sine wave data. The splines index has the largest
squint angle and is the smoothest. The skinny index has low smoothness
and smallest squint angle. As dimension increases the squint angle gets
smaller but smoothness doesn't change much.}

@@

To use JSO for PPGT in the \texttt{tourr} package, specify
\texttt{search\_f\ =\ search\_jellyfish} in the guided tour.

The animation function \texttt{animate\_*()} won't directly render the
animated projection for JSO, as with other existing optimizers since
multiple tour paths will need to be generated. To visualize the path
visited by each individual jellyfish, assign the animation to an object
(\texttt{res\ \textless{}-\ animate\_xy(...)}). This will save the bases
visited by JSO, along with relevant metadata, as a tibble data object
(see Zhang et al. 2021 for more details on the data object). The bases
visited for individual jellyfish can then be extracted and viewed using
the \texttt{planned\_tour()} function.

>>>>

It is expected that for a PP index with high squint angle, the
optimization \eqref{eq-optimization} should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation, shown in the
simulated trace from different index functions in
Figure~\ref{fig-squintability-ill}, suggests that the half-point index
value improvement percentage provides an appropriate mathematical
definition of squintability, which matches the intuition behind this
concept, while being amenable to numerical computation.

@@

To solve the optimization problem embedded in the PPGT with JSO, a
starting projection, an index function, the number of jellyfish, and the
maximum number of iterations are provided as input. Then, the current
projection is evaluated by the index function. The projection is then
moved in a direction determined by a random time control function,
\(c_t\), whose value decreases as the number of trials increases
(subject to randomness) to guide the exploration and exploitation phases
of the algorithm. When \(c_t \ge 0.5\), new directions will be taken
like a jellyfish explores with the ocean trend, defined as the
difference of the current best jellyfish and average of all the current
jellyfish. When \(1- c_t\) is smaller than a randomly generated number
in \([0, 1]\), jellyfish will move passively with small random
perturbations, otherwise, it will move actively towards or away from
another jellyfish based on their index values. A new projection is
accepted if it is an improvement compared to the current one, rejected
otherwise. This process continues and iteratively improves the
projection, until the pre-specified maximum number of trials is reached.

\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, opacityback=0, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, colback=white, leftrule=.75mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, arc=.35mm, coltitle=black, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm]

@@

\begin{quote}
Define the ocean current trend as the difference of the
\texttt{current\_best} and the average of the
\texttt{current\_projections}
\end{quote}

\begin{quote}
Update each projection towards the ocean trend using a random factor and
orthonormalisation
\end{quote}

@@

\caption{\label{fig-success-rate}How success rate is calculated,
illustrated using the optimal projections from 50 optimizations of 8D
pipe data, optimised by CRS, sorted by index value. The pipe shape is
recognizable in the projection index values between 0.933-0.969. Of the
50 simulations, 43 achieved an index value within 0.05 of the best,
resulting in a success rate of 0.86.}

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of iteration) on the success rate is summarized in
Figure~\ref{fig-proportion}. Bootstrap resampling is used to quantify
the uncertainty of the success rate by drawing 500 bootstrap samples
with replacement from the original 50 simulation results for each task:
optimizing a high-dimensional data (d = 4, 6, 8, 10, or 12) with
different number of jellyfish ( 20, 50, or 100) and maximum number of
iteration (50 or 100). As the number of jellyfish and maximum iteration
allowed increase, the success rate also increases. For problems in lower
dimensional (4, 6) search spaces, small parameter values (20 jellyfish
and 50 iterations) are sufficient to achieve a high success rate. Many
more jellyfish (100 jellyfish and 100 iterations) are needed for finding
the pipe structure in d = 10, and 12. This suggests that even JSO will
struggle with high dimensions, and require substantially more
computational time to have an acceptable success rate, especially for
difficult PP indexes based on scagnostics.

@@

\caption{\label{tbl-smoothness-squintability}The columns \(\nu\) and
\(\varsigma\) show the smoothness and squintability measures for three
PP indexes for the sine wave data. The splines index has the largest
squint angle and is the smoothest. The skinny index has low smoothness
and smallest squint angle. As dimension increases the smoothness becomes
smaller but squintability doesn't change much.}

@@

To use JSO for PPGT in the \texttt{tourr} package, specify
\texttt{search\_f\ =\ search\_jellyfish} in the guided tour. The
animation function \texttt{animate\_*()} won't directly render the
animated projection for JSO, as with other existing optimizers since
multiple tour paths will need to be generated. To visualize the path
visited by each individual jellyfish, assign the animation to an object
(\texttt{res\ \textless{}-\ animate\_xy(...)}). This will save all the
bases visited by JSO, along with relevant metadata, as a tibble data
object (see Zhang et al. 2021 for more details on the data object).
Users can then view the path of individual jellyfish through extracting
the relevant bases recorded and construct it as a planned tour (using
the \texttt{planned\_tour()} function). Users may wish to view the
optimisation path of the most successful jellyfish, or a random
selection of jellyfish, or some specific runs that find local maximum.
